Countering language attrition with PanLex and the Web of Data

### 2014-02-24 panlex-rdf-paper/2012/MLOD_PanLex/paper_v5.tex
added v5 of the paper tex file

\begin{abstract}
At present, there are approximately 7,000 living languages in the world.
However, some forecasters expect the world to lose most of this linguistic diversity.
The mission of the PanLex project is to investigate whether it is feasible to make all languages intertranslatable and thus usable for world-wide communication, and, if so, whether that new situation will interfere with this expected language attrition.
This mission entails, among other prerequisites, making the lexicons of the world's languages intertranslatable. To accomplish that, the project is systematically integrating and making publicly available all known lexical translations.
Semantic Web technologies can enhance the power of PanLex by flexibly representing and reasoning with the content of its database and interlinking it with
%linguistic and
other resources and annotations.
%Using Semantic Web technologies can support achieving this goal, especially due
% to inference capabilities and the interlinking of the PanLex data with other data sources.
Conversely, PanLex, with its unique collection of translation links between more than a billion pairs of lexemes from more than 9,000 language varieties, can improve the coverage of the Linguistic Web of Data.
In this dataset description paper we detail how we transformed the content of the PanLex database to RDF,
established conformance with the lemon and GOLD data models,
interlinked it with Lexvo and DBpedia, and published it as Linked Data and via
SPARQL.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
At present, there are about 7,000 living languages in the world\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}.
%\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}. As of 25 January 2014, this list contained 6,975 living individual languages.}
%Nonetheless, processes such as urbanization, nation-state consolidation, and globalization appear to be producing language attrition, with, %for example, the extinction or near-extinction in the last 60 years of from 10\% to over 75\% (depending on the world's region) of all living languages\cite{lang_crisis}.
Nonetheless, processes such as urbanization, nation-state consolidation, and globalization appear to be producing language attrition: For example, in the last 60 years and depending on the world's region, from 10\% to over 75\% of all living languages have become extinct or near-extinct~\cite{lang_crisis}.
Theorists of biolinguistic diversity argue that the loss of language diversity, the loss of human biological knowledge, and the loss of species diversity are mutually supportive and thus that language preservation and revitalization are essential to the preservation of biological diversity~\cite{nettle}.

The mission of the PanLex project is to investigate the possibility that language attrition would be decreased or halted if utterances in any language were efficiently translatable into all other languages, and, as a result, the speakers of all languages could use their languages for world-wide communication.
One prerequisite for panlingual translatablity of discourses is the panlingual translatability of lexicons.
%In fact, lexical translation is almost the entirety of translation in some contexts (such as personal profiles, bibliographic subject headings, content tagging, product price lists, and web navigation controls).
PanLex is designed to support that component by systematically integrating millions of lexical translations, documented in thousands of sources, into a database accessible to the research community, application developers, and the public.
%The database content can be interpreted as a graph linking millions of lexemes with one another in billions of ``is-a-translation-of'' relations, and permitting intelligent automated inference to trillions more such relations that have not been attested.
%It documents the known lexical translations (translations of lexemes) among all
%languages.

%Precisely because the citation forms of lexemes, which is what PanLex directly enables the translation of, are commonly used as labels in collections of data, it is reasonable to speculate that PanLex might facilitate cross-linguistic access to knowledge.

The Semantic Web technologies
standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
lay the foundations for the \emph{Web of Data}: an infrastructure for
%machine-readable and
machine-interpretable data on the Web.

% the basis for the \emph{Web of Data}: a
%: an infrastructure for
%machine-readable
%machine-interpretable data on the Web.

%supporting a machine-readable and machine-interpretable Linked Data
%network.
%This has led to the concept and reality of a \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
Currently there is a growing community leveraging
%Semantic Web
these
technologies for linguistic knowledge which is giving rise to the \emph{Linguistic Linked Open Data (LLOD) cloud}.

% and thereby building a Web of Linguistic Data, also known as the \emph{Linguistic Linked Open Data (LLOD) cloud}.

Our steps to connect PanLex to this Linked Data network are as follows.
In \autoref{sec:triplificate} we introduce the PanLex dataset, present our
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and finally, in \autoref{sec:conclusion}, we conclude our approach and outline future work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
In this section, we first provide an analysis of the PanLex dataset.
Subsequently, we introduce our URI and vocabulary design, which closely
resembles PanLex's original conceptual model.
%, which was developed independently from other ontology engineering approaches.
Afterwards, we briefly describe how we classified PanLex's
instance data using additional data models. Finally, we explain the steps
taken to transform the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
At the core of the PanLex \emph{project}, there is the PanLex \emph{database}, which is created by editors who consult thousands of lexical sources, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri.
The list of sources that have been and are being consulted is available online\footnote{\url{http://panlex.org/tech/plrefs.shtml}}.
The data based on these sources include single- and multi-word expressions, meanings assigned to them, and related information. Far from being mere copies of the sources, which would not make PanLex data interoperable, the data constitute editors' \emph{interpretations} of various sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
% PW: I commented the following sentence out since I removed the detailed number of instances table (language varieties identified/with data; source being consulted/already consulted. I used the bigger numbers and in this sentence the smaller numbers are used, which might confuse
%As of January 2014, the database contained about 20 million meanings and 20 million expressions in about 9,200 language varieties based on the consultation of about 1,500 sources by about 20 editors.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema} and are explained in more detail below.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema_new.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

\todo{The combination of user + source document, which has approved any assertion, is the assertion’s source.}
\begin{itemize}
  \item The starting point of data acquisition is the \emph{source} entity: An editor consults some mono- or multilingual source as mentioned above, interprets it as making assertions about lexical translations, and adds schema-compliant representations of these assertions to the PanLex dataset.
  \item \emph{Expressions} are lexical entities, each defined as a string of characters in some variety of some language. As a first approximation, expressions can be considered lemmas, i.e. dictionary-entry headwords. They differ from traditional headwords, however, in at least two ways. (1) Homographs, such as the verb ``hide'' (conceal) and the noun ``hide'' (animal skin) in English, are treated as a single expression in PanLex. (2) Multiword expressions, such as ``fall in love'', traditionally found in an entry headed by one of their words, such as ``fall'' or ``love'', are treated as independent expressions in PanLex. Expressions are strings of Unicode codepoints subjected to Normalization Form NFC, but this still makes it difficult to search precisely for expressions, so to assist in approximate searching PanLex also adds for each expression a degraded form permitting only lower-case letters and digits.
    \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes, and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
    \item \emph {Language varieties} are the fundamental linguistic entities in PanLex. Expressions are in them, not in languages. The language-variety concept is fine-grained and permissive: there can be many varieties of a language, and editors can add varieties as needed. A language variety is identified with some language code, combined with an arbitrary integer differentiating it from other varieties of the same language. For example, six dialects of Ahtna are identified as ``aht-000'' through ``aht-005''. These symbolic 3-letter, 3-digit labels are, themselves, treated as a (controlled) language variety, so editors can clarify their referents by translating them into natural-language expressions and into expressions in other controlled language varieties, such as the IETF standard BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}}. Such controlled language varieties may be defined as prohibiting synonymy and/or ambiguity.
    \item \emph{Meanings} in PanLex are entities with which expressions are semantically classified. An editor documents that two or more expressions are asserted to be translations or synonyms of one another by assigning a shared meaning to them. For example, depending on an editor's judgment, a source's translation of the German lexeme ``klingen'' into a sequence of alternative English lexemes ``ring, sound, seem'' can be documented by either (1) the assignment of all four expressions to the same meaning or (2) the assignment of two or three meanings to ``klingen'' and of one of those meanings to each of the English expressions. For practicality, PanLex treats expressions and meanings asymmetrically. Newly added translations automatically re-use existing matching expressions, but they don't re-use existing meanings, because there is no straightforward test of equivalence for meanings. PanLex generates new meanings as needed, and meanings are source-specific, leaving it for subsequent research to investigate which meanings should be considered equivalent and thus consolidated. Meanings can further have properties of three types. (1) \emph{Definitions} are descriptions of a meaning, consisting of arbitrary text strings annotated as being in particular language varieties. (2) \emph{Domains} are expressions (e.g., ``medicine'') that classify a meaning, but do not express it. (3) \emph{Meaning identifiers} are strings acting as references to identifiers used in a source.
    \item \emph{Denotations} are assignments of meanings to expressions. Each denotation, in addition to its meaning-expression pair, may be assigned one or more \emph{word classes} (a closed set based on OLIF, the Open Lexicon Interchange Format) and/or \emph{metadata} (arbitrary strings paired as keys and values).
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} are the entities that may act as editors by defining sources and then defining data attributed to those sources, as well as defining language varieties.
  \item Sources in PanLex are defined with about two dozen properties, among them being \emph{licenses}. As of January 2014 eleven different license categories were recognized. They were \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (meaning that one has to ask the author of the resource), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (meaning that there is a copyright claim without a more liberal license also being offered), \emph{PanLex Use Permission} (where an author or publisher has given a source to PanLex for its use), \emph{other}, and \emph{unknown}.
    The distribution of licenses among sources that had been or were being consulted in January 2014 is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

@@

An overview of the number of instances per entity as of January 2014 is given in~\autoref{fig:plx-entity-counts}.
%\begin{table}
%  \centering\begin{scriptsize}
%  \begin{tabular}{lrclr}
%    Entity             & Instances  \\
%    \cline{1-2}
%    Denotations        & 54,278,860 \\  % SELECT COUNT(dn) FROM dn;
%    Languages          &      7,843 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
%    Language Varieties identified           &      9,310 \\  % SELECT COUNT(lv) FROM lv;
%    Language Varieties with data           &      9,239 \\  % 
%    Meanings           & 20,773,371 \\  % SELECT COUNT(mn) FROM mn;
%    Sources being consulted         &      4,190 \\  % 
%    Sources already consulted         &      1,453 \\  % 
%    Expressions        & 19,790,453 \\ % SELECT COUNT(ex) FROM ex;
%    Definitions        &  2,747,892 \\ % SELECT COUNT(df) FROM df;
%    Users              &          23 \\ % SELECT COUNT(us) FROM us;
%  \end{tabular}
%  \end{scriptsize}
%  \caption{Number of instances of main entities in the PanLex database}
%  \label{fig:plx-entity-counts}
%\end{table}

%    Users              &         23 \\ % SELECT COUNT(us) FROM us;
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  & & Entity    & Instances \\
    \cline{1-2}\cline{4-5}
    Denotations        & 54,278,860 & & Languages & 7,843     \\
    Meanings           & 20,773,371 & & Sources   & 4,190     \\
    Expressions        & 19,790,453 & & Users     &    23     \\
    Definitions        &  2,747,892 & & Licenses  &    11     \\
    Language Varieties &      9,310 & & & \\
  \end{tabular}
  \end{scriptsize}
  \caption{Number of instances of main entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations of the schema described in the previous section serve as the base for the development of the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \emph{<http://ld.panlex.org/plx/>}, abbreviated with \emph{plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows.
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \emph{plx:Expression}.
    Their original and degraded textual representations become the values of the properties \emph{rdfs:label} and \emph{plx:degradedText}, respectively.
    Their corresponding language variety is stated using \emph{plx:languageVariety}.
  \item For language and language varieties the classes \emph{plx:Language} and \emph{plx:LanguageVariety} are introduced.
    ISO 639-1 and ISO 639-3 codes become instances of the classes \emph{plx:Iso639-1Code} and \emph{plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \emph{plx:Meaning}.
    Entities of this class may have an identifier assigned with the \emph{plx:identifier} property pointing to an \emph{xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \emph{plx:Definition} class, giving a textual representation (\emph{rdfs:label}) in a certain language variety (\emph{plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \emph{plx:Denotation} class pointing to meanings and expressions via the properties \emph{plx:denotationMeaning} and \emph{plx:\-denotation\-Expression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \emph{plx:wordClass} property pointing to a \emph{plx:WordClass} entity.
  \item All sources share the \emph{plx:Source} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \emph{dc:title} to assign the title of a source, \emph{dc:creator} to give an \emph{xsd:string} containing the author's name.
    At present, we support the different license categories recognized in the database by creating resources of the \emph{plx:License} class.
\end{itemize}

\begin{table}
  \centering
  \begin{scriptsize}
  \fontsize{5.8pt}{5pt}\selectfont
  \begin{tabular}{p{41pt}p{150pt}}
    Class                 & Properties \\
    \toprule
%    \emph{plx:Source} & \emph{plx:registrationDate},\,\emph{rdfs:label},\,\emph{dc:title},\,\emph{dc:creator},\,\emph{dc:quality}, \emph{plx:license},\,\emph{dc:date},\,\emph{foaf:homepage},\,\emph{dc:publisher},\,\emph{dbpedia-owl:isbn} \\
    \emph{plx:Source} & \emph{plx:registrationDate},\,\emph{rdfs:label},\,\emph{dc:title},\,\emph{dc:publisher},\,\emph{dc:quality}, \emph{plx:license},\,\emph{dc:date},\,\emph{foaf:homepage},\,\emph{dc:creator},\,\mbox{\emph{dbpedia-owl:isbn}} \\
    \midrule
    \emph{plx:Language} & \emph{plx:iso639-3Code}, \emph{plx:iso639-1Code} \\
    \midrule
    \emph{plx:LanguageVariety}
                          & \emph{plx:languageVarietyOf}, \emph{rdfs:label} \\
    \midrule
    \mbox{\emph{plx:Iso639-1Code}} & \\
    \midrule
    \mbox{\emph{plx:Iso639-3Code}} & \\
    \midrule
    \emph{plx:Expression}
                          & \emph{plx:languageVariety}, \emph{plx:degradedText}, \emph{rdfs:label} \\
    \midrule
    \emph{plx:Meaning}
                          & \emph{plx:source}, \emph{plx:identifier}, \emph{plx:meaningDefinition} \\
    \midrule
    \emph{plx:Definition}
                          & \emph{plx:languageVariety}, \emph{rdfs:label} \\
    \midrule
    \emph{plx:Denotation}
                          & \emph{plx:denotationMeaning},\,\emph{plx:denotationExpression},\,\emph{plx:wordClass} \\
    \midrule
    \emph{plx:WordClass}  
                          & \emph{rdfs:label} \\
    \midrule
    \emph{plx:License}    & \emph{rdfs:label} \\
    \bottomrule\vspace{3pt}
  \end{tabular}
  \end{scriptsize}
  \caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \emph{rdf:type} properties are omitted for brevity.}
  \label{tbl:vocabulary}
\end{table}

@@

\begin{table}
  \centering\begin{scriptsize}
  \fontsize{6.5pt}{6.4pt}\selectfont
  \begin{tabular}{p{48px}p{62px}p{70px}}
    Panlex                  & lemon                       & GOLD \\
    \midrule
    \emph{plx:Denotation} & --                          & \emph{gold:LinguisticSign} \\
    \emph{plx:Meaning}    & \emph{lemon:LexicalSense} & \emph{gold:SemanticUnit} \\
    \emph{plx:Expression} & \emph{lemon:LexicalEntry} & \emph{gold:FormUnit} \\
  \end{tabular}
  \end{scriptsize}
  \caption{Classes considered to be similar across the re-used vocabulary}
  \label{tbl:sameclasses}
\end{table}

\subsection{Vocabulary Reuse}
\label{sec:vocabulary-reuse}
The PanLex vocabulary is based on PanLex's conceptual schema and
enables all of PanLex's data to be directly exposed as RDF.
%Apart from modeling a PanLex RDF vocabulary
Additionally, we also re-use existing vocabularies, namely
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} as well as the \emph{General Ontology for Linguistic Description} (GOLD)~\cite{farr2003}.
Since these models differ from the PanLex one to some extent, we follow an
incremental approach of aligning the PanLex data with them.
%accordingly. 
% one differ to some
%extend, not the whole schemas were implemented.
\autoref{tbl:sameclasses} shows PanLex classes with their current counterparts
in lemon and GOLD respectively.
The parts implemented in our RDF conversion are displayed in \autoref{fig:goldlemonimplemented}.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{images/pdf/gold_lemon_implemented.pdf}
  \caption{Parts of the GOLD (left) and lemon model (right) re-used in PanLex (URI prefixes are omitted for brevity)}
  \label{fig:goldlemonimplemented}
\end{figure}

>>>>

\begin{abstract}
The world is losing some of its 7,000 languages.
Hypothesizing that language attrition might subside if all languages were inter\-trans\-latable, the PanLex project supports panlingual lexical translation by integrating all known lexical translations.
Semantic Web technologies can flexibly represent and reason with the content of its database and interlink it with linguistic and other resources and annotations.
%Using Semantic Web technologies can support achieving this goal, especially due
% to inference capabilities and the interlinking of the PanLex data with other data sources.
Conversely, PanLex, with its collection of translation links between more than a billion pairs of lexemes from more than 9,000 language varieties, can improve the coverage of the Linguistic Web of Data.
We detail how we transformed the content of the PanLex database to RDF, established conformance with the lemon and GOLD data models,
interlinked it with Lexvo and DBpedia, and published it as Linked Data and via SPARQL.
\end{abstract}

@@

\section{Introduction}
\label{sec:intro}
There are about 7,000 living languages\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}, but language attrition has extinguished or threatened from 10\% to over 75\% of all languages in the last 60 years in various regions\cite{lang_crisis}.
This attrition arguably imperils human biological knowledge and species diversity\cite{nettle}. Hypothetically, panlingual intertranslatability would make all languages more useful and incentivize their preservation and revitalization.

The PanLex project is making all languages' lexicons intertranslatable. In some contexts (e.g., profiles, catalogs, tags, search, and web navigation), lexical translation can be most of the translation load. PanLex systematically integrates lexical translations, found in diverse sources, into a database for research, applications, and public use. The content can be interpreted as a graph linking lexemes in ``is-a-translation-of'' relations, and permitting automated inference to additional, unattested relations.
%It documents the known lexical translations (translations of lexemes) among all
%languages.

The Semantic Web Initiative has led to a stack of technologies standardized by the
\emph{World Wide Web Consortium}\footnote{\url{http://www.w3.org/standards/semanticweb/}}
supporting a machine-readable and \mbox{-i}nter\-pretable Linked Data
network. This has led to a \emph{Web of Data}.
%Technologies that support this aim are already available and in use in the so
%called \emph{Web of Data}.
A growing community is leveraging Semantic Web technologies for linguistic knowledge, building a Web of Linguistic
Data, or a \emph{Linguistic Linked Open Data (LLOD) cloud}.

When linked open data are labeled in any language with lexemes in citation form (lemmas), the data are potentially linkable to, and through, PanLex. Here we describe how we connected PanLex to the Linked Data network.
In \autoref{sec:triplificate} we introduce the dataset, present a
PanLex RDF vocabulary, and explain how we transformed the one into the other and established conformance with additional data models.
\autoref{sec:linking} is about how we linked to other datasets of the LLOD cloud,
and \autoref{sec:publishing} is about the publication of the dataset.
Usage scenarios are given in \autoref{sec:usage}.
In \autoref{sec:related} we discuss related work, and, finally, in \autoref{sec:conclusion}, we summarize our recent and contemplated work.

\section{Triplification of the Raw Data}
\label{sec:triplificate}
Here we analyze the PanLex dataset; introduce our URI and vocabulary design, resembling PanLex's conceptual model;
%, which was developed independently from other ontology engineering approaches.
summarize how we classified PanLex's instance data with additional data models; and explain our transformation of the data to RDF.

@@

\subsection{Analysis of the Original Dataset}
\label{sec:analysis}
The PanLex \emph{database} is created by editors who consult sources\footnote{\url{http://panlex.org/tech/plrefs.shtml}}, such as mono- and multilingual dictionaries, glossaries, standards, and thesauri. The data include single- and multi-word expressions, meanings assigned to them, and related information. While copies of the sources would not be interoperable, PanLex data constitute editors' \emph{interpretations} of sources' \emph{assertions} that two or more \emph{expressions} share a \emph{meaning}\footnote{\url{http://panlex.org/tech/doc/design/panlex-db-design.pdf}}.
The most important entities and relations of PanLex's conceptual model are depicted in \autoref{fig:db-schema}.

\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth]{images/schema-new.png}
  \caption{The PanLex database schema}
  \label{fig:db-schema}
\end{figure}

\begin{itemize}
  \item The \emph{source} entity is the authority to which an editor attributes assertions about lexical translations.
  \item \emph{Expressions} are lexical entities, each uniquely identified with a text, i.e. a string of (Unicode) characters, and a variety of a language. Expressions resemble lemmas or dictionary-entry headwords, but differ from them in at least two ways. (1) Homographs, such as the verb ``hide'' (conceal) and the noun ``hide'' (animal skin) in English, are treated as a single expression in PanLex. (2) Multiword expressions, such as ``fall in love'', traditionally found in an entry headed by one of their words, such as ``fall'' or ``love'', are treated as independent expressions in PanLex. Expression texts are constrained to comply with Unicode Normalization Form NFC.
    \item \emph{Languages} in PanLex are identified using ISO 639-3\footnote{\url{http://www.sil.org/iso639-3/codes.asp}} individual and macrolanguage codes, ISO 639-2\footnote{\url{http://www.loc.gov/standards/iso639-2/}} collective codes, and ISO 639-5\footnote{\url{http://www.loc.gov/standards/iso639-5/}} codes.
    \item \emph {Language varieties} are collections of expressions. Each has a unique identifier: a language code and a distinguishing integer. For example, six dialects of Ahtna are identified as ``aht-000'' through ``aht-005''. These labels are, themselves, treated as a (controlled) language variety, whose expressions (i.e. the labels) are translated into natural languages and other controlled languages (such as the IETF standard BCP 47\footnote{\url{http://tools.ietf.org/html/bcp47}}. The only properties of this entity are ambiguity and synonymy: Controlled language varieties are usually defined to prohibit both of these.
    \item \emph{Meanings} are the entities assigned to expressions, thereby identifying expressions as translations or synonyms. For example, a source's translation of the German expression ``klingen'' into English as ``ring, sound, seem'' can be interpreted as (1) the assignment of a single meaning to all four expressions or (2) the assignment of two or three meanings to ``klingen'' and of one of those to each of the English expressions. Meanings are source-specific. The identification and consolidation of equivalent meanings of distinct sources is a research topic, not a database feature. Meanings can have properties of three types. (1) \emph{Definitions} are descriptions of a meaning, consisting of text strings annotated as being in particular language varieties. (2) \emph{Domains} are expressions (e.g., ``medicine'') that characterize a meaning, but do not express it. (3) \emph{Meaning identifiers} are strings acting as references to identifiers in a source.
    \item \emph{Denotations} are assignments of meanings to expressions. A denotation may have one or more \emph{word classes} (a closed set based on OLIF, the Open Lexicon Interchange Format) and/or \emph{metadata} (arbitrary strings paired as keys and values).
    %For example, \emph{fall} can be a verb or a noun for autumn.
    %Homonyms are those expressions that are connected to multiple meanings.
  \item \emph{Users} may define sources, attribute data to them, and define language varieties.
  \item Among the properties of sources are \emph{licenses}. Some of the license categories are \emph{public domain}, \emph{Creative Commons (CC)}, \emph{request} (author invites inquiries), \emph{GNU General Public License (GPL)}, \emph{GNU Lesser General Public License (LGPL)}, \emph{GNU Free Documentation License (FDL)}, \emph{MIT License}, \emph{copyright} (without a more liberal license), and \emph{PanLex Use Permission} (specific permission for use in PanLex).
    The distribution of licenses as of January 2014 is shown in \autoref{tbl:plx-license-counts}.
\end{itemize}

@@

\autoref{fig:plx-entity-counts} gives entity counts as of January 2014.
\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{lrclr}
    Entity             & Instances  \\
    \cline{1-2}
    Denotations        & 54,278,860 \\  % SELECT COUNT(dn) FROM dn;
    Languages          &      7,843 \\ % SELECT COUNT(lc) FROM lc; -- all language codes (ISO 639-3, ISO 639-2, ISO 639-5) known in the PanLex db
    Language Varieties identified           &      9,310 \\  % SELECT COUNT(lv) FROM lv;
    Language Varieties with data           &      9,239 \\  % 
    Meanings           & 20,773,371 \\  % SELECT COUNT(mn) FROM mn;
    Sources being consulted         &      4,190 \\  % 
    Sources already consulted         &      1,453 \\  % 
    Expressions        & 19,790,453 \\ % SELECT COUNT(ex) FROM ex;
    Definitions        &  2,747,892 \\ % SELECT COUNT(df) FROM df;
    Users              &          23 \\ % SELECT COUNT(us) FROM us;
  \end{tabular}
  \end{scriptsize}
  \caption{Number of instances of main entities in the PanLex database}
  \label{fig:plx-entity-counts}
\end{table}

\subsection{The PanLex vocabulary}
\label{sec:vocabulary}
The entities and relations described above are the base for the PanLex RDF vocabulary.
In general, all PanLex RDF resources reside in the namespace \texttt{\small <http://ld.panlex.org/plx/>}, abbreviated with \texttt{\small plx}.
An example of the resulting ontology is depicted in \autoref{fig:vocabulary} and summarized as follows.
Unless otherwise noted, the URIs of instances of PanLex classes follow the pattern \emph{plx:\{className\}/\{id\}}, where \{className\} is spelled in lower camel case and the \{id\} is the primary key of the corresponding database table.

\begin{itemize}
  \item Expressions are modeled as instances of the class \texttt{\small plx:Expression}.
    Their original and degraded textual representations become the values of the properties \texttt{\small rdfs:label} and \texttt{\small plx:degradedText}, respectively.
    Their corresponding language variety is stated using \texttt{\small plx:languageVariety}.
  \item For language and language varieties the classes \texttt{\small plx:Language} and \texttt{\small plx:LanguageVariety} are introduced.
    \emph{ISO 639-1} and \emph{ISO 639-3 codes} become instances of the classes \texttt{\small plx:Iso639-1Code} and \texttt{\small plx:Iso639-3Code}.
  \item The RDF analog of the PanLex \emph{meaning} is the \texttt{\small plx:Meaning}.
    Entities of this class may have an identifier assigned with the \texttt{\small plx:identifier} property pointing to an \texttt{\small xsd:string} literal.
    Meanings may also have \emph{definitions}, entities of the \texttt{\small plx:Definition} class, giving a textual representation (\texttt{\small rdfs:label}) in a certain language variety (\texttt{\small plx:languageVariety}).
  \item Following the semantics of the PanLex database, meanings and expressions are linked via \emph{denotations}.
    These are entities of the \texttt{\small plx:Denotation} class pointing to meanings and expressions via the properties \texttt{\small plx:denotationMeaning} and \\ \texttt{\small plx:denotationExpression}.
    Denotations may also have a word class assigned to them.
    This can be achieved with the denotation's \texttt{\small plx:wordClass} property pointing to a \texttt{\small plx:WordClass} entity.
  \item All sources share the \texttt{\small plx:Approver} class.
    The characteristics of a source are described using mainly triples with literal objects.
    These are for example \texttt{\small dc:title} to assign the title of a source, \texttt{\small dc:creator} to give an \texttt{\small xsd:string} containing the author's name. At present, we support the different license categories recognized in the database by creating resources of the \texttt{\small plx:License} class.
\end{itemize}

\begin{table}
  \begin{tiny}
  \begin{tabular}{p{52px}p{140px}}
    Class                 & Properties \\
    \toprule
    \texttt{plx:Approver} & \mbox{\texttt{plx:registrationDate}, \texttt{rdfs:label}, \texttt{dc:title},}
                            \mbox{\texttt{dc:creator}, \texttt{plx:license}, \texttt{dc:date}, \texttt{plx:quality},}
                            \mbox{\texttt{foaf:homepage}, \texttt{dc:publisher}, \texttt{dbpedia-owl:isbn}} \\
    \midrule
    \texttt{plx:Language} & \texttt{plx:iso639-3Code}, \texttt{plx:iso639-1Code} \\
    \midrule
    \texttt{plx:LanguageVariety}
                          & \texttt{plx:languageVarietyOf}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Iso639-1Code} & \\
    \midrule
    \texttt{plx:Iso639-3Code} & \\
    \midrule
    \texttt{plx:Expression}
                          & \mbox{\texttt{plx:languageVariety}, \texttt{plx:degradedText},} \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Meaning}
                          & \mbox{\texttt{plx:approver}, \texttt{plx:identifier},} \texttt{plx:meaningDefinition} \\
    \midrule
    \texttt{plx:Definition}
                          & \texttt{plx:languageVariety}, \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:Denotation}
                          & \texttt{plx:denotationMeaning}, \texttt{plx:denotationExpression}, \texttt{plx:wordClass} \\
    \midrule
    \texttt{plx:WordClass}  
                          & \texttt{rdfs:label} \\
    \midrule
    \texttt{plx:License}  & \texttt{rdfs:label} \\
    \bottomrule
  \end{tabular}
  \end{tiny}
  \caption{Classes and properties used in the PanLex RDF vocabulary. Note that all \texttt{rdf:type} properties are omitted for brevity.}
  \label{tbl:vocabulary}
\end{table}

@@

\begin{table}
  \centering\begin{scriptsize}
  \begin{tabular}{p{48px}p{62px}p{70px}}
    Panlex                  & lemon                       & GOLD \\
    \midrule
    \texttt{plx:Denotation} & --                          & \texttt{gold:LinguisticSign} \\
    \texttt{plx:Meaning}    & \texttt{lemon:LexicalSense} & \texttt{gold:SemanticUnit} \\
    \texttt{plx:Expression} & \texttt{lemon:LexicalEntry} & \texttt{gold:FormUnit} \\
  \end{tabular}
  \end{scriptsize}
  \caption{Classes considered to be similar across the re-used vocabulary}
  \label{tbl:sameclasses}
\end{table}

\subsection{Vocabulary Reuse}
\label{sec:vocabulary-reuse}
The PanLex vocabulary is based on PanLex's conceptual schema and
enables all of PanLex's data to be directly exposed as RDF.
%Apart from modeling a PanLex RDF vocabulary
Additionally, we also re-use existing vocuabularies, namely
the \emph{Lexicon Model for Ontologies} (lemon)~\cite{lemon2011} as well as the \emph{General Ontology for Linguistic Description} (GOLD)~\cite{farr2003}.
Since these models differ from the PanLex one to some extent, we follow an
incremental approach of aligning the PanLex data with them.
%accordingly. 
% one differ to some
%extend, not the whole schemas were implemented.
\autoref{tbl:sameclasses} shows PanLex classes with their current counterparts
in lemon and GOLD respectively.
The parts implemented in our RDF conversion are displayed in \autoref{fig:goldlemonimplemented}.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{images/pdf/gold_lemon_implemented.pdf}
  \caption{Parts of the GOLD (left) and lemon model (right) re-used in PanLex (URI prefixes are omitted for brevity)}
  \label{fig:goldlemonimplemented}
\end{figure}
